---
title: "Research Ethics in CSS and Web Data Collection"
author: "Sebastian Stier"
lesson: 4
institute: University of Mannheim & GESIS -- Leibniz Institute for the Social Sciences
date: "2026-03-04"
date-format: "YYYY-MM-DD"
bibliography: references.bib
footer: "[Computational Social Science and Digital Behavioral Data, University of Mannheim](https://sebstier.github.io/ma_css26/)"
format: 
  fakegesis-revealjs: 
    code-line-numbers: false

---


##

```{r, echo=FALSE, warning=FALSE, out.width="105%", message=FALSE}
#| html-table-processing: none

library(openxlsx)
library(tidyverse)
library(gt)
library(gtExtras)
read.xlsx("../material/data/schedule_in_slides.xlsx") %>%
    rename(`Required reading` = "Required.reading") %>%
    head(7) %>% 
    gt() %>%
    tab_header(md("**Seminar dates and topics**")) %>%
    #tab_header("**Seminar dates and topics**") %>%
    #fmt_markdown() %>% #columns = TRUE
    # cols_width(Date ~ px(400)#,
    #            #Topics ~ px(350)#,
    #            #`Required reading` ~ px(400)
    #            ) %>%
    cols_width(Date ~ pct(20),
               Topics ~ pct(30),
               `Required reading` ~ pct(50)
               ) %>%
    tab_options(data_row.padding = px(3)) %>%
    tab_options(heading.title.font.size = 14,
                column_labels.font.weight = "bold",
                table.font.size = 12) %>%
     gt_highlight_rows(
     rows = 4,
     fill = "lightblue"
   ) %>%
  fmt_markdown() %>% 
  cols_align(
    align = "left",
    columns = everything()
  )


```


## Agenda for today


1. *R* session: Data wrangling and getting started with text data

2. Research ethics


## Leftover from last class


```{r ai_disclosure, echo=FALSE, warning=FALSE, out.width="105%", message=FALSE}
#| html-table-processing: none

library(openxlsx)
library(tidyverse)
library(gt)
library(gtExtras)
read.xlsx("../material/data/ai_disclosure.xlsx") %>%
    rename(`AI system used` = Al.used) %>%
    gt() %>%
    tab_header(md("**AI disclosure statement, examples**")) %>%
    #tab_header("**Seminar dates and topics**") %>%
    #fmt_markdown() %>% #columns = TRUE
    # cols_width(Date ~ px(400)#,
    #            #Topics ~ px(350)#,
    #            #`Required reading` ~ px(400)
    #            ) %>%
   #  cols_width(Date ~ pct(20),
   #             Topics ~ pct(30),
   #             `Required reading` ~ pct(50)
   #             ) %>%
   #  tab_options(data_row.padding = px(3)) %>%
    tab_options(heading.title.font.size = 23,
                column_labels.font.weight = "bold",
                table.font.size = 20) %>%
   #   gt_highlight_rows(
   #   rows = 4,
   #   fill = "lightblue"
   # ) %>%
  fmt_markdown() %>% 
  cols_align(
    align = "left",
    columns = everything()
  )


```


# 1. *R* session: Data wrangling and getting started with text data {background-color="#58748F"}

## Trump Twitter Archive


![](../material/img/trump_twitter){width=100%}


::: {style="font-size: 40%;"}
[https://www.thetrumparchive.com](https://www.thetrumparchive.com)
:::


## Coding
- Download file **2_data_wrangling.R** from [https://sebastianstier.com/ma_css26/material.html](https://sebastianstier.com/ma_css25/material.html)
- Open it in RStudio, ideally still within your project



# 2. Research ethics {background-color="#58748F"}


## Research ethics?

:::{.incremental}

- Moral principles and actions guiding and shaping research

    - from inception to completion,
    - through dissemination and sharing of findings,
    - including archiving and future use.
:::

. . .

:::{.incremental}


- Research ethics in the social sciences
    - Initially "patient protection" model of medical research
    - Today broader scope including consideration of benefits, risks and harms to all persons connected with the research
    - Including social responsibilities of researchers

:::


## Who are the relevant actors?

. . . 

![](../material/img/actors_data_protection){width=100%}


## Data protection


- Data protection
    - Part of fundamental right to privacy (or "informational freedom")
- "Privacy is a personal condition of life characterised by seclusion from, and therefore absence of acquaintance by, the public" [@neethling2005personality: 233]
- Prevention of unwanted disclosure of personal information or the misuse of such information


## Data protection: legal foundations

- Legal framework in the European Union
    - Charter of Fundamental Rights of the EU (Art. 8)
    - General Data Protection Regulation (GDPR)
    - National and sub-national data protection acts
    - Specialized laws
- Some legal documents and commentary
    - [Digital Services Act data access for researchers](https://algorithmic-transparency.ec.europa.eu/news/faqs-dsa-data-access-researchers-2023-12-13_en)
    - [German law on text and data mining](https://dejure.org/gesetze/UrhG/60d.html)



## Informed consent

![](../material/img/informed_consent){fig-align="center"}

- Bridge between data protection and research ethics
- Informed consent means for example:
    - information
    - transparency
    - minimal requirement $\rightarrow$ chance *not* to consent
- Status quo in social media research: lack of informed consent



## Social networking sites

::: {style="font-size: 93%;"}

"We define social network sites as web-based services that allow individuals to 

1. construct a public or semi-public profile within a bounded system, 
2. articulate a list of other users with whom they share a connection, and 
3. view and traverse their list of connections and those made by others within the system. 

The nature and nomenclature of these connections may vary from site to site."
[@boyd.2007: 211]

$\rightarrow$ **widely differing platform affordances**

:::


## Are all these researchers doing something wrong?


:::: {.columns}

::: {.column width="60%"}


![](../material/img/report_SM_platforms){width=100%}

::: {style="font-size: 40%;"}
Lukito, J., et al. (2023). [The State of Digital Media Data Research, 2023. Media and Democracy Data Cooperative.](https://mddatacoop.org/files/2023/State%20of%20Digital%20Media%20Data%20Research%202023.pdf)
:::

:::


::: {.column width="40%"}

![](../material/img/number_twitter_studies){width=90%}

::: {style="font-size: 40%;"}

Yu, J., & Muñoz-Justicia, J. (2020). A Bibliometric Overview of Twitter-Related Studies Indexed in Web of Science. [*Future Internet*, 12(5), 91.](https://doi.org/10.3390/fi12050091) 
:::

:::


::::


## Think-pair-share

1. Discuss the literature for today with your neighbor (5 min.)
2. Share+discuss your position on **trade offs between research ethics in CSS and data access restrictions** with the full class

:::: {.columns}

::: {.column width="50%"}


![](../material/img/fiesler){width=90%}

::: {style="font-size: 40%;"}
Fiesler, C., & Proferes, N. (2018). ['Participant' Perceptions of Twitter Research Ethics.*Social Media + Society*, 4(1).](https://doi.org/10.1177/2056305118763366)

:::

:::

::: {.column width="50%"}

![](../material/img/freelon){width=100%}

::: {style="font-size: 40%;"}

Freelon, D. (2018). [Computational Research in the Post-API Age. *Political Communication*, 35(4), 665–668.](https://osf.io/preprints/socarxiv/56f4q)


:::

:::


::::



## Legitimate academic interests


There are legitimate academic interests for studying online platforms!

. . . 

- Understanding the impact of digital media on society and democracy
- Keeping online platforms accountable
- Mitigating "systemic risks" (see [https://digital-strategy.ec.europa.eu/en/consultations/guidelines-providers-very-large-online-platforms-and-very-large-online-search-engines-mitigation](European Union, 2024))
- Preserving historical material
- ... and more reasons?


## What can we do to ensure the integrity of our CSS research?

. . . 

::: {style="font-size: 88%;"}
- Awareness: incorporate ethical considerations when designing and conducting our research
- Define ethical "contextual factors" [@fiesler_participant_2018: 9]

    - "justify what and how many tweets, what the study is about, who is conducting it, what methods researchers use"
- Keep the DBD safe and protected
- Analyze DBD at the aggregated group-/variable-level
- Avoid unnecessary "forensic analysis" of individual data patterns / users
- Not publishing/printing any personal information in a paper/thesis
- "Rewarding" participants with monetary incentives or information

:::



# Thank you for your attention! See you on 18 March 2026 {background-color="#58748F"} 

## References
